Well-quasi-orderings on word languages

### 2025-10-10 src/infixes.tex
lemma 10 polishing

    Assume that $w$ is \kl{ultimately uniformly recurrent}. Consider a sequence
    of words $\seqof{w_i}[i \in \Nat]$ that are finite \kl{infixes} of $w$. Because $w$ is
    \kl{ultimately uniformly recurrent}, there exists a bound $N_0$ such that
    $w_{\geq N_0}$ is \kl{uniformly recurrent}. Let $i < N_0$, we claim that,
    without loss of generality, only finitely many words in the sequence
    $\seqof{w_i}[i \in \Nat]$ can be found starting at the position $i$ in $w$. Indeed, if
    it is not the case, then we have an infinite subsequence of words that are
    all comparable for the \kl{infix relation}, and therefore a \kl{good
    sequence}, because the \kl{infix relation} is \kl{well-founded}. We can
    therefore assume that all words in the sequence $\seqof{w_i}[i \in \Nat]$ are such that
    they start at a position $i \geq N_0$. But then, they are all finite
    \kl{infixes} of $w_{\geq N_0}$, which is a \kl{uniformly recurrent} word,
    whose set of finite \kl{infixes} is \kl{well-quasi-ordered}
    (\cref{uniformly-recurrent:lem}).

    Conversely, assume that the set of finite infixes of $w$ is
    \kl{well-quasi-ordered}. Let us write $\mathsf{Rec}(w)$ the set of finite
    \kl{infixes} of $w$ that appear infinitely often. We can similarly define
    $\mathsf{Rec}(w_{\geq i})$ for any (infinite) suffix of $w$. The sequence
    $R_i \defined \mathsf{Rec}(w_{\geq i})$ is a sequence of \kl{downwards
    closed} sets of finite words, included in the set of finite infixes of $w$
    by definition. Because the latter is \kl{well-quasi-ordered}, there exists
    an $N_0 \in \Nat$, such that $\bigcap_{i \in \Nat} R_i = R_{N_0}$.
    Now, consider $v \defined w_{\geq N_0}$. By construction, 
    every finite infix of $v$ 
    appears infinitely often in $v$. Let us consider 
    some finite infix $u \infleq v$, we claim that there is a bound $N_u$
    on the distance
    between two consecutive occurrences of $u$ in $v$.
    Indeed, if it is not the case, then there exists an infinite sequence
    $\seqof{u x_i u}[i \in \Nat]$ of infixes of $v$, such that $x_i$ is a word that is
    of size $\geq i$.
    Because the finite infixes of $w$ (hence, of $v$) are \kl{well-quasi-ordered},
    one can extract an infinite set of indices $I \subseteq \Nat$
    such that $u x_i u \infleq u x_{j} u$ for all $i \leq j \in I$.
    In particular, $u x_i u \infleq u x_{j} u$ for some $j > |x_i|$, 
    which contradicts the fact that $u x_j u$ coded two consecutive
    occurrences of $u$ in $v$.

>>>>

    Assume that $w$ is \kl{ultimately uniformly recurrent}. Consider a sequence
    of words $\seqof{w_i}[i \in \Nat]$ that are finite \kl{infixes} of $w$. Because $w$ is
    \kl{ultimately uniformly recurrent}, there exists a bound $N_0$ such that
    $w_{\geq N_0}$ is \kl{uniformly recurrent}. Let $i < N_0$, we claim that,
    without loss of generality, only finitely many words in the sequence
    $\seqof{w_i}[i \in \Nat]$ can be found starting at the position $i$ in $w$. Indeed, if
    it is not the case, then we have an infinite subsequence of words that are
    all comparable for the \kl{infix relation}, and therefore a \kl{good
    sequence}, because the \kl{infix relation} is \kl{well-founded}. We can
    therefore assume that all words in the sequence $\seqof{w_i}[i \in \Nat]$ are such that
    they start at a position $i \geq N_0$. But then they are all finite
    \kl{infixes} of $w_{\geq N_0}$, which is a \kl{uniformly recurrent} word,
    whose set of finite \kl{infixes} is \kl{well-quasi-ordered}
    (\cref{uniformly-recurrent:lem}).

    Conversely, assume that the set of finite infixes of $w$ is
    \kl{well-quasi-ordered}. Let us write $\mathsf{Rec}(w)$ the set of finite
    \kl{infixes} of $w$ that appear infinitely often. We can similarly define
    $\mathsf{Rec}(w_{\geq i})$ for any (infinite) suffix of $w$. The sequence
    $R_i \defined \mathsf{Rec}(w_{\geq i})$ is a descending sequence of \kl{downwards
    closed} sets of finite words, included in the set of finite infixes of $w$
    by definition. Because the latter is \kl{well-quasi-ordered}, there exists
    an $N_0 \in \Nat$, such that $\bigcap_{i \in \Nat} R_i = R_{N_0}$.
    Now, consider $v \defined w_{\geq N_0}$. By construction, 
    every finite infix of $v$ 
    appears infinitely often in $v$. 
    Given
    some finite infix $u \infleq v$, we 
    there exists a bound $N_u$
    on the distance
    between two consecutive occurrences of $u$ in $v$.
    Indeed, if it is not the case, then there exists an infinite sequence
    $\seqof{u x_i u}[i \in \Nat]$ of infixes of $v$, such that $x_i$ is a word
    of size $\geq i$
    and no shorter word $u y u$ is an infix of $u x_i u$.
    Because the finite infixes of $w$ (hence, of $v$) are \kl{well-quasi-ordered},
    one can extract an infinite set of indices $I \subseteq \Nat$
    such that $u x_i u \infleq u x_{j} u$ for all $i \leq j \in I$.
    In particular, $u x_i u \infleq u x_{j} u$ for some $j > |x_i|$, 
    which contradicts the fact that $u x_j u$ coded two consecutive
    occurrences of $u$ in $v$.
